---
title: "Project proposal"
author: "Jack, Margaret, Ben"
output: github_document
---

```{r load-packages, message = FALSE}
library(tidyverse)
library(broom)
library(skimr)
library(jsonlite)
```

## 1. Introduction

Our general research question is: How do energy usage trends in New York City public schools vary on a per-student basis, and what geographic and socioeconomic factors are associated with those variations?

The primary motivation is to identify energy inefficiency hotspots and to explore potential environmental equity issues. We will also estimate carbon costs based on the type and amount of energy used by schools and compare these across neighborhoods of differing financial health.

## 2. Data

We placed our raw data files in the /data folder and created a README in that folder that includes variable definitions and data dimensions.


```{r Main Dataset}
library(jsonlite)
library(dplyr)
library(skimr)

data <- fromJSON("../data/dataset.json") %>%
  as_tibble()
glimpse(data)
skim(data)

```


```{r Finance Dataset}
library(readr)
library(dplyr)
library(skimr)

wealth <- read_csv("../data/Neighborhood_Financial_Health_Digital_Mapping_and_Data_Tool.csv")
glimpse(wealth)
skim(wealth)
```

```{r Enrollment Dataset}
library(readr)
library(dplyr)
library(skimr)

enrollment <- read_csv("../data/Enrollment.csv")
glimpse(enrollment)
skim(enrollment)
```
## 3. Data analysis plan

Our analysis will explore patterns of energy use per student across New York City public schools, linking them with neighborhood financial health indicators.

We will visualize and quantify how energy consumption and estimated carbon emissions differ by borough and neighborhood wealth. 

We included two sample graphs that we may end up using.

**Planned analyses**

1. Normalization: Compute energy per student for each school.

2. Carbon estimation: Apply EPA emission factors by energy type (electricity, gas, steam).

3. Geographic comparisons: Aggregate by borough and neighborhood.

4. Socioeconomic correlation: Compare energy intensity with income and financial-health indicators.

5. Outlier detection: Identify schools with unusually high per-student usage.

**Preliminary exploratory analysis**

```{r message=FALSE}
# Total energy usage per borough

library(tidyverse)
library(viridis)

energy <- as_tibble(data)

energy_type <- "Total Usage (mmBTUs)"

energy_filtered <- energy |>
  filter(measurement == energy_type)

usage_cols <- energy_filtered |>
  select(
    -building_address,
    -borough,
    -measurement,
    -postcode,
    -latitude,
    -longitude,
    -community_board,
    -council_district,
    -census_tract,
    -bin,
    -bbl,
    -nta
  ) |>
  colnames()

energy_numeric <- energy_filtered |>
  mutate(across(all_of(usage_cols), ~ suppressWarnings(as.numeric(.)))) |>
  mutate(total_energy = rowSums(across(all_of(usage_cols)), na.rm = TRUE)) |>
  mutate(
    borough_name = factor(
      borough,
      levels = c("1","2","3","4","5"),
      labels = c("Manhattan","Bronx","Brooklyn","Queens","Staten Island")
    )
  )

borough_totals <- energy_numeric |>
  group_by(borough_name) |>
  summarise(
    total_energy_borough = sum(total_energy, na.rm = TRUE)
  )

ggplot(borough_totals) +
  geom_col(aes(
    x = borough_name,
    y = total_energy_borough,
    fill = borough_name
  )) +
  scale_fill_viridis_d(option = "D", name = "Borough Name") + 
  labs(
    title = paste0(energy_type, " by Borough"),
    x = "Borough",
    y = energy_type
  ) +
  theme_minimal() + theme(legend.position = "none")

```

```{r}
library(tidyverse)
library(readxl)
library(viridis)

energy <- as_tibble(data)

demo <- read_excel("../data/enrollment-public-district-2024-25-all-students.xlsx",
                   sheet = "public-district-2024-25")

nyc_dists <- demo |>
  filter(
    County %in% c("BRONX", "KINGS", "NEW YORK", "QUEENS", "RICHMOND"),
    `Subgroup Name` == "All Students"
  ) |>
  mutate(
    borough = case_when(
      County == "NEW YORK" ~ "Manhattan",
      County == "BRONX" ~ "Bronx",
      County == "KINGS" ~ "Brooklyn",
      County == "QUEENS" ~ "Queens",
      County == "RICHMOND" ~ "Staten Island",
      TRUE ~ NA_character_
    ),
    `PreK-12 Total` = readr::parse_number(`PreK-12 Total`)
  ) |>
  filter(!is.na(borough)) |>
  select(borough, `PreK-12 Total`)

enroll_borough <- nyc_dists |>
  group_by(borough) |>
  summarise(total_students = sum(`PreK-12 Total`, na.rm = TRUE), .groups = "drop")

energy_type <- "Total Usage (mmBTUs)"

energy_filtered <- energy |>
  filter(measurement == energy_type)

usage_cols <- energy_filtered |>
  select(
    -building_address, -borough, -measurement, -postcode, -latitude,
    -longitude, -community_board, -council_district, -census_tract,
    -bin, -bbl, -nta
  ) |>
  colnames()

energy_numeric <- energy_filtered |>
  mutate(across(all_of(usage_cols), ~ suppressWarnings(as.numeric(.)))) |>
  mutate(
    total_energy = rowSums(across(all_of(usage_cols)), na.rm = TRUE),
    borough = factor(
      borough,
      levels = c("1","2","3","4","5"),
      labels = c("Manhattan","Bronx","Brooklyn","Queens","Staten Island")
    )
  )

energy_borough <- energy_numeric |>
  group_by(borough) |>
  summarise(total_energy = sum(total_energy, na.rm = TRUE), .groups = "drop")

per_capita <- energy_borough |>
  inner_join(enroll_borough, by = "borough") |>
  mutate(energy_per_student = total_energy / total_students)

ggplot(per_capita) +
  geom_col(aes(
    x = fct_reorder(borough, energy_per_student),
    y = energy_per_student,
    fill = borough
  )) +
  scale_fill_viridis_d(option = "D") +
  labs(
    title = "Energy Per Student by Borough",
    x = "Borough",
    y = "mmBTUs per student"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```



```{r energy-boxplot, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)

energy_box <- energy %>%
  filter(grepl("Total Usage", measurement, ignore.case = TRUE)) %>%
  mutate(across(starts_with(c("aug", "sep", "oct", "nov", "dec","jan", "feb", "mar", "apr", "may", "jun")), as.numeric)) %>%
  rowwise() %>%
  mutate(avg_monthly_usage = mean(c_across(starts_with(c("aug", "sep", "oct", "nov", "dec","jan", "feb", "mar", "apr", "may", "jun"))), na.rm = TRUE)) %>%
  ungroup()

ggplot(energy_box, aes(x = borough, y = avg_monthly_usage)) +
  geom_boxplot(fill = "darkorange", alpha = 0.7) +
  labs(
    title = "Distribution of Average Monthly Energy Usage by Borough",
    x = "Borough",
    y = "Average Monthly Usage (MMBtu)"
  ) +
  theme_minimal()
ggsave(
  "energy_boxplot.png",
  plot = get_last_plot(),
  device = NULL,
  path = NULL,
  scale = 1,
  width = NA,
  height = NA,
  units = c("in", "cm", "mm", "px"),
  dpi = 300,
  limitsize = TRUE,
  bg = NULL,
  create.dir = FALSE,
)
```


```{r}
library(tidyverse)
library(viridis)

# Clean Borough and ensure income is numeric
wealth_boro <- df |>
  mutate(
    Borough = str_trim(Borough),
    Median_Income = suppressWarnings(as.numeric(Median_Income))
  ) |>
  filter(!is.na(Borough), !is.na(Median_Income)) |>
  group_by(Borough) |>
  summarise(median_income = median(Median_Income, na.rm = TRUE), .groups = "drop")

# Use same palette as earlier borough plots if available; otherwise create one
pal <- if (exists("pal")) pal else setNames(
  viridis(5, option = "D"),
  c("Manhattan","Bronx","Brooklyn","Queens","Staten Island")
)

ggplot(wealth_boro) +
  geom_col(aes(
    x = fct_reorder(Borough, median_income),
    y = median_income,
    fill = Borough
  )) +
  scale_fill_manual(values = pal) +
  labs(
    title = "Wealth Per Capita (Proxy) by Borough",
    subtitle = "Proxy = Median household income",
    x = "Borough",
    y = "Median household income (USD)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

```{r}
library(tidyverse)
library(viridis)

wealth_df <- df |>
  mutate(Borough = str_trim(Borough),
         Median_Income = suppressWarnings(as.numeric(Median_Income))) |>
  filter(!is.na(Borough), !is.na(Median_Income)) |>
  group_by(Borough) |>
  summarise(median_income = median(Median_Income, na.rm = TRUE), .groups = "drop") |>
  rename(borough = Borough)

compare_df <- per_capita |>
  select(borough, energy_per_student, total_energy, total_students) |>
  inner_join(wealth_df, by = "borough")

boro_order <- per_capita |> arrange(desc(energy_per_student)) |> pull(borough) |> as.character()
pal <- if (exists("pal")) pal else setNames(viridis(5, option = "D"), boro_order)

p_scatter <- ggplot(compare_df) +
  geom_point(aes(x = median_income, y = energy_per_student, color = borough), size = 4) +
  geom_text(aes(x = median_income, y = energy_per_student, label = borough, color = borough),
            vjust = -1, size = 4, show.legend = FALSE) +
  scale_color_manual(values = pal) +
  labs(title = "Energy per Student vs Median Income by Borough",
       x = "Median household income (USD)",
       y = "mmBTUs per student") +
  theme_minimal() + theme(legend.position = "none")

p_scatter
```

