---
title: "Project memo"
author: "Jack, Margaret, Ben"
output: github_document
editor_options: 
  markdown: 
    wrap: 72
---

Design choices: Standardized coloring for burrows to enhance readability
between graphs Virdits colorblind safe for our palette Legends removed
to increase plot size & reduce clutter because categories are already on
axis Per student normalization to control for enrollment differences
Borough mapping via county to increase readability / familiairity Alt
text to encourage accessibility

## Data Clean Up Steps for Overall Data

### Step 1: Import Packages

```{r load-packages}
library(tidyverse)
library(broom)
library(skimr)
library(jsonlite)
```

### Step 2: Import Datasets

```{r Main Dataset}
library(jsonlite)
library(dplyr)
library(skimr)

data <- fromJSON("../data/dataset.json") %>%
  as_tibble()
glimpse(data)
skim(data)

```

```{r Finance Dataset}
library(readr)
library(dplyr)
library(skimr)

wealth <- read_csv("../data/Neighborhood_Financial_Health_Digital_Mapping_and_Data_Tool.csv")
glimpse(wealth)
skim(wealth)
```

```{r Enrollment Dataset}
library(readr)
library(dplyr)
library(skimr)

enrollment <- read_csv("../data/Enrollment.csv")
glimpse(enrollment)
skim(enrollment)
```

## Plots

### Plot 1: Total Usage by Borough

#### Data cleanup steps specific to plot 1

### Step 1: Convert Raw Dataset

```{r dataset-to-tibble}
energy <- as_tibble(data)
```

### Step 2: Filter Energy Dataset

```{r filter-energy}
energy_type <- "Total Usage (mmBTUs)"

energy_filtered <- energy |>
  filter(measurement == energy_type)

```

### Step 3: Select Columns in Energy Dataset

```{r select-energy}
usage_cols <- energy_filtered |>
  select(
    -building_address,
    -borough,
    -measurement,
    -postcode,
    -latitude,
    -longitude,
    -community_board,
    -council_district,
    -census_tract,
    -bin,
    -bbl,
    -nta
  ) |>
  colnames()

```

### Step 4: Convert Columns to Numeric

```{r energy-to-numeric}
energy_numeric <- energy_filtered |>
  mutate(across(all_of(usage_cols), ~ suppressWarnings(as.numeric(.))))

```

### Step 5: Compute Total Energy per School

```{r energy-per-school}
energy_numeric <- energy_numeric |>
  mutate(total_energy = rowSums(across(all_of(usage_cols)), na.rm = TRUE))

```

### Step 6: Borough Numbers to Names

```{r Borough-to-names}
energy_numeric <- energy_numeric |>
  mutate(
    borough_name = factor(
      borough,
      levels = c("1","2","3","4","5"),
      labels = c("Manhattan","Bronx","Brooklyn","Queens","Staten Island")
    )
  )

```

### Step 7: Aggregate by Borough

```{r aggregate-by-borough}
borough_totals <- energy_numeric |>
  group_by(borough_name) |>
  summarise(total_energy_borough = sum(total_energy, na.rm = TRUE))

```

### Final Plot 1

```{r plot-1}
ggplot(borough_totals) +
  geom_col(aes(
    x = borough_name,
    y = total_energy_borough,
    fill = borough_name
  )) +
  scale_fill_viridis_d(option = "D") +
  labs(
    title = paste0(energy_type, " by Borough"),
    x = "Borough",
    y = energy_type
  ) +
  theme_minimal() +
  theme(legend.position = "none")

```

## Plot 2: Per Student Plot

### Step 1: Import and Filter Enrollment Dataset

```{r import-enrollment}
demo <- readxl::read_excel(
  "../data/enrollment-public-district-2024-25-all-students.xlsx",
  sheet = "public-district-2024-25"
)

```

### Step 2: Standardize Borough Names and Parse Enrollment

```{r parse-enrollment}
nyc_dists <- demo |>
  filter(
    County %in% c("BRONX", "KINGS", "NEW YORK", "QUEENS", "RICHMOND"),
    `Subgroup Name` == "All Students"
  ) |>
  mutate(
    borough = case_when(
      County == "NEW YORK" ~ "Manhattan",
      County == "BRONX" ~ "Bronx",
      County == "KINGS" ~ "Brooklyn",
      County == "QUEENS" ~ "Queens",
      County == "RICHMOND" ~ "Staten Island"
    ),
    `PreK-12 Total` = readr::parse_number(`PreK-12 Total`)
  ) |>
  select(borough, `PreK-12 Total`)

```

### Step 3: Aggregate Enrollment Totals by Borough

```{r aggregate-enrollment}
enroll_borough <- nyc_dists |>
  group_by(borough) |>
  summarise(total_students = sum(`PreK-12 Total`, na.rm = TRUE))

```

### Step 4: Aggregate Energy by Borough

```{r aggregate-energy}
energy_borough <- energy_numeric |>
  group_by(borough_name) |>
  summarise(total_energy = sum(total_energy, na.rm = TRUE))

```

### Step 5: Compute Energy per Student

```{r energy-per-student}
per_capita <- energy_borough |>
  inner_join(enroll_borough, by = c("borough_name" = "borough")) |>
  mutate(energy_per_student = total_energy / total_students)
```

### Final Plot 2

```{r plot-2}
ggplot(per_capita) +
  geom_col(aes(
    x = fct_reorder(borough_name, energy_per_student),
    y = energy_per_student,
    fill = borough_name
  )) +
  scale_fill_viridis_d(option = "D") +
  labs(
    title = "Energy Per Student by Borough",
    x = "Borough",
    y = "mmBTUs per student"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

```

## Plot 3: Boxplot

### Step 1: Filter Rows for Total Energy Usage

```{r filter-rows}
energy_box <- energy |>
  filter(grepl("Total Usage", measurement, ignore.case = TRUE))

```

### Step 2: Convert Monthly Columns to Numeric

```{r monthly-to-numeric}
energy_box_numeric <- energy_box |>
  mutate(across(
    starts_with(c("aug","sep","oct","nov","dec","jan","feb","mar","apr","may","jun")),
    ~ suppressWarnings(as.numeric(.))
  ))

```

### Step 3: Get Average Monthly Usage

```{r monthly-average-usage}
energy_box_numeric <- energy_box_numeric |>
  rowwise() |>
  mutate(
    avg_monthly_usage = mean(
      c_across(starts_with(c("aug","sep","oct","nov","dec","jan","feb","mar","apr","may","jun"))),
      na.rm = TRUE
    )
  ) |>
  ungroup()

```

### Step 4: Compute Borough Medians

```{r medians}
wealth_boro <- wealth |>
  mutate(
    Borough = stringr::str_trim(Borough),
    Median_Income = suppressWarnings(as.numeric(Median_Income))
  ) |>
  filter(!is.na(Borough), !is.na(Median_Income)) |>
  group_by(Borough) |>
  summarise(median_income = median(Median_Income, na.rm = TRUE))

```

### Final Plot 3

```{r plot-3}
ggplot(energy_box_numeric, aes(
  x = borough,
  y = avg_monthly_usage
)) +
  geom_boxplot(fill = "darkorange", alpha = 0.7) +
  labs(
    title = "Distribution of Average Monthly Energy Usage by Borough",
    x = "Borough",
    y = "Average Monthly Usage (MMBtu)"
  ) +
  theme_minimal()

```

### Final Plot 4

```{r}
wealth_source <- if (exists("df")) {
  df
} else {
  wealth_file <- list.files("../data", pattern = "(?i)Neighborhood_Financial_Health.*\\.csv$", full.names = TRUE)
  readr::read_csv(wealth_file[1], show_col_types = FALSE)
}

bcols <- names(wealth_source)
borough_col <- bcols[grepl("^borough$", bcols, ignore.case = TRUE) | grepl("^boro(ugh)?s?$", bcols, ignore.case = TRUE)]
if (length(borough_col) == 0) stop("Could not find a 'Borough' column. Columns are: ", paste(bcols, collapse = ", "))

income_candidates <- c("Median_Income","Median_Household_Income","Income","median_income","median_household_income","income")
income_col <- bcols[match(tolower(income_candidates), tolower(bcols), nomatch = 0)]
if (length(income_col) == 0) stop("Could not find an income column. Looked for: ", paste(income_candidates, collapse = ", "),
                                  ". Columns are: ", paste(bcols, collapse = ", "))
income_col <- income_col[1]

wealth_boro <- wealth_source |>
  mutate(
    borough_std = stringr::str_trim(.data[[borough_col[1]]]),
    income_val  = suppressWarnings(as.numeric(.data[[income_col]]))
  ) |>
  filter(!is.na(borough_std), !is.na(income_val)) |>
  group_by(borough_std) |>
  summarise(median_income = median(income_val, na.rm = TRUE), .groups = "drop") |>
  rename(borough = borough_std)

pal <- if (exists("pal")) pal else setNames(
  viridis(5, option = "D"),
  c("Manhattan","Bronx","Brooklyn","Queens","Staten Island")
)

p4 <- ggplot(wealth_boro) +
  geom_col(aes(x = fct_reorder(borough, median_income), y = median_income, fill = borough)) +
  scale_fill_manual(values = pal) +
  labs(
    title = "Wealth Per Capita (Proxy) by Borough",
    subtitle = "Proxy = Median household income",
    x = "Borough",
    y = "Median household income (USD)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

p4
```

```{r plots-PNG}
p1 <- ggplot(borough_totals) +
  geom_col(aes(
    x = borough_name,
    y = total_energy_borough,
    fill = borough_name
  )) +
  scale_fill_viridis_d(option = "D") +
  labs(
    title = paste0(energy_type, " by Borough"),
    x = "Borough",
    y = energy_type
  ) +
  theme_minimal() +
  theme(legend.position = "none")

p1
ggsave("plot1-total-usage-by-borough.png", p1, width = 6, height = 4, dpi = 300)

p2 <- ggplot(per_capita) +
  geom_col(aes(
    x = fct_reorder(borough_name, energy_per_student),
    y = energy_per_student,
    fill = borough_name
  )) +
  scale_fill_viridis_d(option = "D") +
  labs(
    title = "Energy Per Student by Borough",
    x = "Borough",
    y = "mmBTUs per student"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

p2
ggsave("plot2-energy-per-student.png", p2, width = 6, height = 4, dpi = 300)

p3 <- ggplot(energy_box_numeric, aes(
  x = borough,
  y = avg_monthly_usage
)) +
  geom_boxplot(fill = "darkorange", alpha = 0.7) +
  labs(
    title = "Distribution of Average Monthly Energy Usage by Borough",
    x = "Borough",
    y = "Average Monthly Usage (MMBtu)"
  ) +
  theme_minimal()

p3
ggsave("plot3-monthly-usage-boxplot.png", p3, width = 6, height = 4, dpi = 300)

p4 <- ggplot(wealth_boro) +
  geom_col(aes(x = fct_reorder(borough, median_income), y = median_income, fill = borough)) +
  scale_fill_manual(values = pal) +
  labs(
    title = "Wealth Per Capita (Proxy) by Borough",
    subtitle = "Proxy = Median household income",
    x = "Borough",
    y = "Median household income (USD)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

p4
ggsave("plot4-wealth-per-capita-by-borough.png", p4, width = 6, height = 4, dpi = 300)
```

###Final Plot 5

### Step 1: Define Conversion Units
```{r create-units}
# Energy → MMBtu
kwh_to_mmbtu      <- 0.003412
mlb_to_mmbtu      <- 1.19
therm_to_mmbtu    <- 0.1
gal_no2_to_mmbtu  <- 0.1387
gal_no4_to_mmbtu  <- 0.145
gal_no56_to_mmbtu <- 0.150

# MMBtu → kg CO₂
kg_per_mmbtu_gas   <- 53.06
kg_per_mmbtu_oil2  <- 73.16
kg_per_mmbtu_oil4  <- 75.29
kg_per_mmbtu_oil56 <- 78.80
kg_per_kwh_grid    <- 0.25
```

### Step 2: Identify Numeric Usage Columns
```{r id-columns}
usage_cols_all <- energy |>
  select(-building_address, -borough, -measurement, -postcode, -latitude,
         -longitude, -community_board, -council_district, -census_tract,
         -bin, -bbl, -nta) |>
  colnames()

```

### Step 3: Convert to Numeric
```{r to-numeric}
carbon_ready <- energy |>
  mutate(
    across(all_of(usage_cols_all), ~ suppressWarnings(as.numeric(.))),
    total_raw = rowSums(across(all_of(usage_cols_all)), na.rm = TRUE)
  )

```

### Step 4: Classify Fuel Types
```{r fuel-types}
carbon_ready <- carbon_ready |>
  mutate(
    fuel = case_when(
      str_detect(measurement, "(?i)electric") ~ "electricity",
      str_detect(measurement, "(?i)steam")    ~ "steam",
      str_detect(measurement, "(?i)gas|therm")~ "natural_gas",
      str_detect(measurement, "(?i)oil.*2")   ~ "oil_2",
      str_detect(measurement, "(?i)oil.*4")   ~ "oil_4",
      str_detect(measurement, "(?i)oil.*(5|6)") ~ "oil_56",
      str_detect(measurement, "(?i)total usage") ~ "total_mmbtu",
      TRUE ~ "other"
    ),
    borough_name = factor(
      borough,
      levels = c("1","2","3","4","5"),
      labels = c("Manhattan","Bronx","Brooklyn","Queens","Staten Island")
    )
  ) |>
  filter(fuel %in% c("electricity","steam","natural_gas","oil_2","oil_4","oil_56"))

```

### Step 5: Convert Fuel Types to kg CO2
```{r convert-fuel-types}
carbon_by_row <- carbon_ready |>
  mutate(
    kg_co2 = case_when(
      fuel == "electricity" ~ total_raw * kg_per_kwh_grid,

      fuel == "steam"       ~ total_raw *
                                mlb_to_mmbtu *
                                kg_per_mmbtu_gas,

      fuel == "natural_gas" ~ total_raw *
                                therm_to_mmbtu *
                                kg_per_mmbtu_gas,

      fuel == "oil_2"       ~ total_raw *
                                gal_no2_to_mmbtu *
                                kg_per_mmbtu_oil2,

      fuel == "oil_4"       ~ total_raw *
                                gal_no4_to_mmbtu *
                                kg_per_mmbtu_oil4,

      fuel == "oil_56"      ~ total_raw *
                                gal_no56_to_mmbtu *
                                kg_per_mmbtu_oil56,

      TRUE ~ NA_real_
    )
  )

```

### Step 6: Aggregate to Borough Level

```{r aggregate-boroughs}
carbon_borough <- carbon_by_row |>
  group_by(borough_name) |>
  summarise(kg_co2 = sum(kg_co2, na.rm = TRUE), .groups = "drop") |>
  mutate(t_co2 = kg_co2 / 1000)

```

### Step 7: Set Color Palette
```{r colors}
if (!exists("pal")) {
  pal <- setNames(
    viridis(5, option = "D"),
    c("Manhattan","Bronx","Brooklyn","Queens","Staten Island")
  )
}
```

### Final Plot 5
```{r plot-5}
p5 <- ggplot(carbon_borough) +
  geom_col(aes(
    x = fct_reorder(borough_name, t_co2),
    y = t_co2,
    fill = borough_name
  )) +
  scale_fill_manual(values = pal) +
  labs(
    title = "Estimated Carbon Emissions by Borough",
    x = "Borough",
    y = "tCO₂ (estimated)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

p5

```

