---
title: "Project memo"
author: "Jack, Margaret, Ben"
output: github_document
---

This document should contain a detailed account of the data clean up for your data and the design choices you are making for your plots. For instance you will want to document choices you've made that were intentional for your graphic, e.g. color you've chosen for the plot. Think of this document as a code script someone can follow to reproduce the data cleaning steps and graphics in your handout.


## Data Clean Up Steps for Overall Data

### Step 1: Import Packages
```{r load-packages}
library(tidyverse)
library(broom)
library(skimr)
library(jsonlite)
```

### Step 2: Import Datasets
```{r Main Dataset}
library(jsonlite)
library(dplyr)
library(skimr)

data <- fromJSON("../data/dataset.json") %>%
  as_tibble()
glimpse(data)
skim(data)

```
```{r Finance Dataset}
library(readr)
library(dplyr)
library(skimr)

wealth <- read_csv("../data/Neighborhood_Financial_Health_Digital_Mapping_and_Data_Tool.csv")
glimpse(wealth)
skim(wealth)
```
```{r Enrollment Dataset}
library(readr)
library(dplyr)
library(skimr)

enrollment <- read_csv("../data/Enrollment.csv")
glimpse(enrollment)
skim(enrollment)
```


## Plots

### Plot 1: Total Usage by Borough 

#### Data cleanup steps specific to plot 1

### Step 1: Convert Raw Dataset
```{r dataset-to-tibble}
energy <- as_tibble(data)
```

### Step 2: Filter Energy Dataset

```{r filter-energy}
energy_type <- "Total Usage (mmBTUs)"

energy_filtered <- energy |>
  filter(measurement == energy_type)

```

### Step 3: Select Columns in Energy Dataset
```{r select-energy}
usage_cols <- energy_filtered |>
  select(
    -building_address,
    -borough,
    -measurement,
    -postcode,
    -latitude,
    -longitude,
    -community_board,
    -council_district,
    -census_tract,
    -bin,
    -bbl,
    -nta
  ) |>
  colnames()

```

### Step 4: Convert Columns to Numeric

```{r energy-to-numeric}
energy_numeric <- energy_filtered |>
  mutate(across(all_of(usage_cols), ~ suppressWarnings(as.numeric(.))))

```

### Step 5: Compute Total Energy per School

```{r energy-per-school}
energy_numeric <- energy_numeric |>
  mutate(total_energy = rowSums(across(all_of(usage_cols)), na.rm = TRUE))

```

### Step 6: Borough Numbers to Names
```{r Borough-to-names}
energy_numeric <- energy_numeric |>
  mutate(
    borough_name = factor(
      borough,
      levels = c("1","2","3","4","5"),
      labels = c("Manhattan","Bronx","Brooklyn","Queens","Staten Island")
    )
  )

```

### Step 7: Aggregate by Borough
```{r aggregate-by-borough}
borough_totals <- energy_numeric |>
  group_by(borough_name) |>
  summarise(total_energy_borough = sum(total_energy, na.rm = TRUE))

```

#### Final Plot 1

```{r plot-1}
ggplot(borough_totals) +
  geom_col(aes(
    x = borough_name,
    y = total_energy_borough,
    fill = borough_name
  )) +
  scale_fill_viridis_d(option = "D") +
  labs(
    title = paste0(energy_type, " by Borough"),
    x = "Borough",
    y = energy_type
  ) +
  theme_minimal() +
  theme(legend.position = "none")

```


### Plot 2: Per Student Plot
### Step 1: Import and Filter Enrollment Dataset
```{r import-enrollment}
demo <- readxl::read_excel(
  "../data/enrollment-public-district-2024-25-all-students.xlsx",
  sheet = "public-district-2024-25"
)

```

### Step 2: Standardize Borough Names and Parse Enrollment

```{r parse-enrollment}
nyc_dists <- demo |>
  filter(
    County %in% c("BRONX", "KINGS", "NEW YORK", "QUEENS", "RICHMOND"),
    `Subgroup Name` == "All Students"
  ) |>
  mutate(
    borough = case_when(
      County == "NEW YORK" ~ "Manhattan",
      County == "BRONX" ~ "Bronx",
      County == "KINGS" ~ "Brooklyn",
      County == "QUEENS" ~ "Queens",
      County == "RICHMOND" ~ "Staten Island"
    ),
    `PreK-12 Total` = readr::parse_number(`PreK-12 Total`)
  ) |>
  select(borough, `PreK-12 Total`)

```

### Step 3: Aggregate Enrollment Totals by Borough
```{r aggregate-enrollment}
enroll_borough <- nyc_dists |>
  group_by(borough) |>
  summarise(total_students = sum(`PreK-12 Total`, na.rm = TRUE))

```

### Step 4: Aggregate Energy by Borough 
```{r aggregate-energy}
energy_borough <- energy_numeric |>
  group_by(borough_name) |>
  summarise(total_energy = sum(total_energy, na.rm = TRUE))

```

### Step 5: Compute Energy per Student
```{r energy-per-student}
per_capita <- energy_borough |>
  inner_join(enroll_borough, by = c("borough_name" = "borough")) |>
  mutate(energy_per_student = total_energy / total_students)
```

#### Final Plot 2
```{r plot-2}
ggplot(per_capita) +
  geom_col(aes(
    x = fct_reorder(borough_name, energy_per_student),
    y = energy_per_student,
    fill = borough_name
  )) +
  scale_fill_viridis_d(option = "D") +
  labs(
    title = "Energy Per Student by Borough",
    x = "Borough",
    y = "mmBTUs per student"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

```

### Plot 3: Boxplot

### Step 1: Filter Rows for Total Energy Usage
```{r filter-rows}
energy_box <- energy |>
  filter(grepl("Total Usage", measurement, ignore.case = TRUE))

```

### Step 2: Convert Monthly Columns to Numeric
```{r monthly-to-numeric}
energy_box_numeric <- energy_box |>
  mutate(across(
    starts_with(c("aug","sep","oct","nov","dec","jan","feb","mar","apr","may","jun")),
    ~ suppressWarnings(as.numeric(.))
  ))

```

### Step 3: Get Average Monthly Usage
```{r monthly-average-usage}
energy_box_numeric <- energy_box_numeric |>
  rowwise() |>
  mutate(
    avg_monthly_usage = mean(
      c_across(starts_with(c("aug","sep","oct","nov","dec","jan","feb","mar","apr","may","jun"))),
      na.rm = TRUE
    )
  ) |>
  ungroup()

```

### Step 4: Compute Borough Medians
```{r medians}
wealth_boro <- wealth |>
  mutate(
    Borough = stringr::str_trim(Borough),
    Median_Income = suppressWarnings(as.numeric(Median_Income))
  ) |>
  filter(!is.na(Borough), !is.na(Median_Income)) |>
  group_by(Borough) |>
  summarise(median_income = median(Median_Income, na.rm = TRUE))

```
#### Final Plot 3
```{r plot-3}
ggplot(energy_box_numeric, aes(
  x = borough,
  y = avg_monthly_usage
)) +
  geom_boxplot(fill = "darkorange", alpha = 0.7) +
  labs(
    title = "Distribution of Average Monthly Energy Usage by Borough",
    x = "Borough",
    y = "Average Monthly Usage (MMBtu)"
  ) +
  theme_minimal()

```