---
title: "Project memo"
author: "Jack, Margaret, Ben"
output: github_document
editor_options: 
  markdown: 
    wrap: 72
---

Design choices: Standardized coloring for burrows to enhance readability
between graphs Virdits colorblind safe for our palette Legends removed
to increase plot size & reduce clutter because categories are already on
axis Per student normalization to control for enrollment differences
Borough mapping via county to increase readability / familiairity Alt
text to encourage accessibility

## Data Clean Up Steps for Overall Data

### Step 1: Import Packages

```{r load-packages}
library(tidyverse)
library(broom)
library(skimr)
library(jsonlite)
```

### Step 2: Import Datasets

```{r Main Dataset}
library(jsonlite)
library(dplyr)
library(skimr)

data <- fromJSON("../data/dataset.json") %>%
  as_tibble()
glimpse(data)
skim(data)

```

```{r Finance Dataset}
library(readr)
library(dplyr)
library(skimr)

wealth <- read_csv("../data/Neighborhood_Financial_Health_Digital_Mapping_and_Data_Tool.csv")
glimpse(wealth)
skim(wealth)
```

```{r Enrollment Dataset}
library(readr)
library(dplyr)
library(skimr)

enrollment <- read_csv("../data/Enrollment.csv")
glimpse(enrollment)
skim(enrollment)
```

## Plots

### Plot 1: Total Usage by Borough

### Step 1: Convert Raw Dataset

```{r dataset-to-tibble}
energy <- as_tibble(data)
```

### Step 2: Filter Energy Dataset

```{r filter-energy}
energy_type <- "Total Usage (mmBTUs)"

energy_filtered <- energy |>
  filter(measurement == energy_type)

```

### Step 3: Select Columns in Energy Dataset

```{r select-energy}
usage_cols <- energy_filtered |>
  select(
    -building_address,
    -borough,
    -measurement,
    -postcode,
    -latitude,
    -longitude,
    -community_board,
    -council_district,
    -census_tract,
    -bin,
    -bbl,
    -nta
  ) |>
  colnames()

```

### Step 4: Convert Columns to Numeric

```{r energy-to-numeric}
energy_numeric <- energy_filtered |>
  mutate(across(all_of(usage_cols), ~ suppressWarnings(as.numeric(.))))

```

### Step 5: Compute Total Energy per School

```{r energy-per-school}
energy_numeric <- energy_numeric |>
  mutate(total_energy = rowSums(across(all_of(usage_cols)), na.rm = TRUE))

```

### Step 6: Borough Numbers to Names

```{r Borough-to-names}
energy_numeric <- energy_numeric |>
  mutate(
    borough_name = factor(
      borough,
      levels = c("1","2","3","4","5"),
      labels = c("Manhattan","Bronx","Brooklyn","Queens","Staten Island")
    )
  )

```

### Step 7: Aggregate by Borough

```{r aggregate-by-borough}
borough_totals <- energy_numeric |>
  group_by(borough_name) |>
  summarise(total_energy_borough = sum(total_energy, na.rm = TRUE))

```

### Final Plot 1

```{r plot-1, fig.alt = "Vertical bar chart showing total annual energy usage (in mmBTUs) for five NYC boroughs: Manhattan, Bronx, Brooklyn, Queens, and Staten Island. Manhattan has the highest total usage, followed by the Bronx and Brooklyn, while Queens and Staten Island are much lower."}
ggplot(borough_totals) +
  geom_col(aes(
    x = borough_name,
    y = total_energy_borough,
    fill = borough_name
  )) +
  scale_fill_viridis_d(option = "D") +
  labs(
    title = paste0(energy_type, " by Borough"),
    x = "Borough",
    y = energy_type
  ) +
  theme_minimal() +
  theme(legend.position = "none")

```

## Plot 2: Per Student Plot

### Step 1: Import and Filter Enrollment Dataset

```{r import-enrollment}
demo <- readxl::read_excel(
  "../data/enrollment-public-district-2024-25-all-students.xlsx",
  sheet = "public-district-2024-25"
)

```

### Step 2: Standardize Borough Names and Parse Enrollment

```{r parse-enrollment}
nyc_dists <- demo |>
  filter(
    County %in% c("BRONX", "KINGS", "NEW YORK", "QUEENS", "RICHMOND"),
    `Subgroup Name` == "All Students"
  ) |>
  mutate(
    borough = case_when(
      County == "NEW YORK" ~ "Manhattan",
      County == "BRONX" ~ "Bronx",
      County == "KINGS" ~ "Brooklyn",
      County == "QUEENS" ~ "Queens",
      County == "RICHMOND" ~ "Staten Island"
    ),
    `PreK-12 Total` = readr::parse_number(`PreK-12 Total`)
  ) |>
  select(borough, `PreK-12 Total`)

```

### Step 3: Aggregate Enrollment Totals by Borough

```{r aggregate-enrollment}
enroll_borough <- nyc_dists |>
  group_by(borough) |>
  summarise(total_students = sum(`PreK-12 Total`, na.rm = TRUE))

```

### Step 4: Aggregate Energy by Borough

```{r aggregate-energy}
energy_borough <- energy_numeric |>
  group_by(borough_name) |>
  summarise(total_energy = sum(total_energy, na.rm = TRUE))

```

### Step 5: Compute Energy per Student

```{r energy-per-student}
per_capita <- energy_borough |>
  inner_join(enroll_borough, by = c("borough_name" = "borough")) |>
  mutate(energy_per_student = total_energy / total_students)
```

### Final Plot 2

```{r plot-2, fig.alt = "Vertical bar chart showing energy usage per student (in mmBTUs) across Manhattan, Staten Island, Bronx, Brooklyn, and Queens. Manhattan and Staten Island have the highest per-student usage, the Bronx and Brooklyn fall in the middle, and Queens has the lowest."}
ggplot(per_capita) +
  geom_col(aes(
    x = fct_reorder(borough_name, energy_per_student),
    y = energy_per_student,
    fill = borough_name
  )) +
  scale_fill_viridis_d(option = "D") +
  labs(
    title = "Energy Per Student by Borough",
    x = "Borough",
    y = "mmBTUs per student",
  ) +
  theme_minimal() +
  theme(legend.position = "none")

```

## Plot 3: Boxplot

### Step 1: Filter Rows for Total Energy Usage

```{r filter-rows}
energy_box <- energy |>
  filter(grepl("Total Usage", measurement, ignore.case = TRUE))

```

### Step 2: Convert Monthly Columns to Numeric

```{r monthly-to-numeric}
energy_box_numeric <- energy_box |>
  mutate(across(
    starts_with(c("aug","sep","oct","nov","dec","jan","feb","mar","apr","may","jun")),
    ~ suppressWarnings(as.numeric(.))
  ))

```

### Step 3: Get Average Monthly Usage

```{r monthly-average-usage}
energy_box_numeric <- energy_box_numeric |>
  rowwise() |>
  mutate(
    avg_monthly_usage = mean(
      c_across(starts_with(c("aug","sep","oct","nov","dec","jan","feb","mar","apr","may","jun"))),
      na.rm = TRUE
    )
  ) |>
  ungroup()

```

### Step 4: Compute Borough Medians

```{r medians}
energy_box_numeric <- energy_box_numeric |>
  mutate(
    borough = factor(
      borough,
      levels = c("1","2","3","4","5"),
      labels = c("Manhattan","Bronx","Brooklyn","Queens","Staten Island")
    )
  ) |>
  rowwise() |>
  mutate(
    avg_monthly_usage = mean(
      c_across(starts_with(c("aug","sep","oct","nov","dec","jan","feb","mar","apr","may","jun"))),
      na.rm = TRUE
    )
  ) |>
  ungroup()
```

### Final Plot 3
```{r plot-3, fig.alt = "Box plot comparing average monthly energy usage (in mmBTUs) for schools in Manhattan, Bronx, Brooklyn, Queens, and Staten Island. Manhattan and the Bronx show the greatest spread and highest values, indicating variability in building performance, while Brooklyn and Queens have tighter, more consistent distributions."}
ggplot(energy_box_numeric, aes(
  x = borough,
  y = avg_monthly_usage
)) +
  geom_boxplot(fill = "darkorange", alpha = 0.7) +
  labs(
    title = "Distribution of Average Monthly Energy Usage by Borough",
    x = "Borough",
    y = "Average Monthly Usage (MMBtu)"
  ) +
  theme_minimal()
```

## Plot 4

### Step 1: Load Wealth Dataset

```{r load-wealth-dataset}
wealth_source <- wealth 
```

### Step 2: Clean Borough Names

```{r clean-boroughs}
wealth_boro <- wealth_source |>
  mutate(
    borough_std = stringr::str_trim(Borough),
    borough_std = case_when(
      str_detect(borough_std, "(?i)man")     ~ "Manhattan",
      str_detect(borough_std, "(?i)bronx")   ~ "Bronx",
      str_detect(borough_std, "(?i)brook")   ~ "Brooklyn",
      str_detect(borough_std, "(?i)queen")   ~ "Queens",
      str_detect(borough_std, "(?i)richmond|staten") ~ "Staten Island",
      TRUE ~ borough_std
    )
  )

```

### Step 3: Convert Income to Numeric

```{r convert-to-numeric}
wealth_boro <- wealth_boro |>
  mutate(
    income_val = suppressWarnings(as.numeric(Median_Income))
  ) |>
  filter(!is.na(borough_std), !is.na(income_val))
```

### Step 4: Aggregate Median Income

```{r aggregate}
wealth_boro <- wealth_boro |>
  group_by(borough_std) |>
  summarise(
    median_income = median(income_val, na.rm = TRUE),
    .groups = "drop"
  ) |>
  rename(borough = borough_std)
```

### Step 5: Color Palette
```{r colors}
if (!exists("pal")) {
  pal <- setNames(
    viridis::viridis(5, option = "D"),
    c("Manhattan", "Bronx", "Brooklyn", "Queens", "Staten Island")
  )
}
```

### Final Plot 4

```{r plot-4, fig.alt = "Vertical bar chart displaying median household income for Manhattan, Staten Island, Brooklyn, Queens, and the Bronx. Manhattan has the highest income by a wide margin, Staten Island is next, Brooklyn and Queens fall in the middle, and the Bronx is lowest."}
p4 <- ggplot(wealth_boro) +
  geom_col(aes(
    x = forcats::fct_reorder(borough, median_income),
    y = median_income,
    fill = borough
  )) +
  scale_fill_manual(values = pal) +
  labs(
    title = "Wealth Per Capita (Proxy) by Borough",
    subtitle = "Proxy = Median Household Income",
    x = "Borough",
    y = "Median household income (USD)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

p4
```


###Final Plot 5

### Step 1: Define Conversion Units

```{r create-units}
# Energy → MMBtu
kwh_to_mmbtu      <- 0.003412
mlb_to_mmbtu      <- 1.19
therm_to_mmbtu    <- 0.1
gal_no2_to_mmbtu  <- 0.1387
gal_no4_to_mmbtu  <- 0.145
gal_no56_to_mmbtu <- 0.150

# MMBtu → kg CO₂
kg_per_mmbtu_gas   <- 53.06
kg_per_mmbtu_oil2  <- 73.16
kg_per_mmbtu_oil4  <- 75.29
kg_per_mmbtu_oil56 <- 78.80
kg_per_kwh_grid    <- 0.25
```

### Step 2: Identify Numeric Usage Columns

```{r id-columns}
usage_cols_all <- energy |>
  select(-building_address, -borough, -measurement, -postcode, -latitude,
         -longitude, -community_board, -council_district, -census_tract,
         -bin, -bbl, -nta) |>
  colnames()

```

### Step 3: Convert to Numeric

```{r to-numeric}
carbon_ready <- energy |>
  mutate(
    across(all_of(usage_cols_all), ~ suppressWarnings(as.numeric(.))),
    total_raw = rowSums(across(all_of(usage_cols_all)), na.rm = TRUE)
  )

```

### Step 4: Classify Fuel Types

```{r fuel-types}
carbon_ready <- carbon_ready |>
  mutate(
    fuel = case_when(
      str_detect(measurement, "(?i)electric") ~ "electricity",
      str_detect(measurement, "(?i)steam")    ~ "steam",
      str_detect(measurement, "(?i)gas|therm")~ "natural_gas",
      str_detect(measurement, "(?i)oil.*2")   ~ "oil_2",
      str_detect(measurement, "(?i)oil.*4")   ~ "oil_4",
      str_detect(measurement, "(?i)oil.*(5|6)") ~ "oil_56",
      str_detect(measurement, "(?i)total usage") ~ "total_mmbtu",
      TRUE ~ "other"
    ),
    borough_name = factor(
      borough,
      levels = c("1","2","3","4","5"),
      labels = c("Manhattan","Bronx","Brooklyn","Queens","Staten Island")
    )
  ) |>
  filter(fuel %in% c("electricity","steam","natural_gas","oil_2","oil_4","oil_56"))

```

### Step 5: Convert Fuel Types to kg CO2

```{r convert-fuel-types}
carbon_by_row <- carbon_ready |>
  mutate(
    kg_co2 = case_when(
      fuel == "electricity" ~ total_raw * kg_per_kwh_grid,

      fuel == "steam"       ~ total_raw *
                                mlb_to_mmbtu *
                                kg_per_mmbtu_gas,

      fuel == "natural_gas" ~ total_raw *
                                therm_to_mmbtu *
                                kg_per_mmbtu_gas,

      fuel == "oil_2"       ~ total_raw *
                                gal_no2_to_mmbtu *
                                kg_per_mmbtu_oil2,

      fuel == "oil_4"       ~ total_raw *
                                gal_no4_to_mmbtu *
                                kg_per_mmbtu_oil4,

      fuel == "oil_56"      ~ total_raw *
                                gal_no56_to_mmbtu *
                                kg_per_mmbtu_oil56,

      TRUE ~ NA_real_
    )
  )

```

### Step 6: Aggregate to Borough Level

```{r aggregate-boroughs}
carbon_borough <- carbon_by_row |>
  group_by(borough_name) |>
  summarise(kg_co2 = sum(kg_co2, na.rm = TRUE), .groups = "drop") |>
  mutate(t_co2 = kg_co2 / 1000)

```

### Step 7: Set Color Palette

```{r colors}
if (!exists("pal")) {
  pal <- setNames(
    viridis(5, option = "D"),
    c("Manhattan","Bronx","Brooklyn","Queens","Staten Island")
  )
}
```

### Final Plot 5

```{r plot-5, fig.alt = "Vertical bar chart showing estimated annual carbon emissions (in metric tons of CO₂) from school buildings across Manhattan, Bronx, Brooklyn, Queens, and Staten Island. Manhattan has the highest total emissions, followed by the Bronx and Brooklyn, while Queens and Staten Island have the lowest."}

p5 <- ggplot(carbon_borough) +
  geom_col(aes(
    x = fct_reorder(borough_name, t_co2),
    y = t_co2,
    fill = borough_name
  )) +
  scale_fill_manual(values = pal) +
  labs(
    title = "Estimated Carbon Emissions by Borough",
    x = "Borough",
    y = "tCO₂ (estimated)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

p5

```

```{r gg-save-plots-png}
p1 <- ggplot(borough_totals) +
  geom_col(aes(x = borough_name, y = total_energy_borough)) +
  labs(
    title = "Total Energy Usage by Borough",
    x = "Borough",
    y = "Total Usage (mmBTUs)"
  ) +
  theme_minimal()

p2 <- ggplot(per_capita) +
  geom_col(aes(x = fct_reorder(borough_name, energy_per_student),
               y = energy_per_student)) +
  labs(
    title = "Energy Per Student by Borough",
    x = "Borough",
    y = "Energy per Student (mmBTUs)"
  ) +
  theme_minimal()

p3 <- ggplot(energy_box_numeric, aes(x = borough, y = avg_monthly_usage)) +
  geom_boxplot() +
  labs(
    title = "Distribution of Average Monthly Energy Usage by Borough",
    x = "Borough",
    y = "Average Monthly Usage (MMBtu)"
  ) +
  theme_minimal()

p4 <- ggplot(wealth_boro) +
  geom_col(aes(x = fct_reorder(borough, median_income),
               y = median_income)) +
  labs(
    title = "Wealth Per Capita by Borough",
    x = "Borough",
    y = "Median Household Income (USD)"
  ) +
  theme_minimal()

p5 <- ggplot(carbon_borough) +
  geom_col(aes(
    x = fct_reorder(borough_name, t_co2),
    y = t_co2,
    fill = borough_name
  )) +
  scale_fill_manual(values = pal) +
  labs(
    title = "Estimated Carbon Emissions by Borough",
    x = "Borough",
    y = "tCO₂ (estimated)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

ggsave("plot1-total-usage.png", p1, width = 6, height = 4, dpi = 300)
ggsave("plot2-energy-per-student.png", p2, width = 6, height = 4, dpi = 300)
ggsave("plot3-monthly-usage.png", p3, width = 6, height = 4, dpi = 300)
ggsave("plot4-wealth.png", p4, width = 6, height = 4, dpi = 300)
ggsave("plot5-carbon.png", p5, width = 6, height = 4, dpi = 300)
```

