---
title: "Project memo"
author: "Jack, Margaret, Ben"
output: github_document
---

This document should contain a detailed account of the data clean up for your data and the design choices you are making for your plots. For instance you will want to document choices you've made that were intentional for your graphic, e.g. color you've chosen for the plot. Think of this document as a code script someone can follow to reproduce the data cleaning steps and graphics in your handout.

```{r load-packages, message = FALSE}
library(tidyverse)
library(broom)
library(tidyverse)
library(broom)
library(skimr)
library(jsonlite)
```

```{r Enrollment Dataset}
library(readr)
library(dplyr)
library(skimr)

enrollment <- read_csv("../data/Enrollment.csv")
glimpse(enrollment)
skim(enrollment)
```
```{r Finance Dataset}
library(readr)
library(dplyr)
library(skimr)

wealth <- read_csv("../data/Neighborhood_Financial_Health_Digital_Mapping_and_Data_Tool.csv")
glimpse(wealth)
skim(wealth)
```
```{r Main Dataset}
library(jsonlite)
library(dplyr)
library(skimr)

data <- fromJSON("../data/dataset.json") %>%
  as_tibble()
glimpse(data)
skim(data)

```

## Data Clean Up Steps for Overall Data

### Step 1: _________

```{r}

```

### Step 2: ________

## Plots

### ggsave example for saving plots

```{r starwars ggsave example}
p1 <- starwars |>
  filter(mass < 1000, 
         species %in% c("Human", "Cerean", "Pau'an", "Droid", "Gungan")) |>
  ggplot() +
  geom_point(aes(x = mass, 
                 y = height, 
                 color = species)) +
  labs(x = "Weight (kg)", 
       y = "Height (m)",
       color = "Species",
       title = "Weight and Height of Select Starwars Species",
       caption = paste("This data comes from the starwars api: https://swapi.py43.com"))


ggsave("example-starwars.png", width = 4, height = 4)

ggsave("example-starwars-wide.png", width = 6, height = 4)
```



### Plot 1: _________

#### Data cleanup steps specific to plot 1

These data cleaning sections are optional and depend on if you have some data cleaning steps specific to a particular plot

```{r}

```


#### Final Plot 1

```{r}

```


### Plot 2: _________
```{r message=FALSE}

library(tidyverse)
library(viridis)

energy <- as_tibble(data)

energy_type <- "Total Usage (mmBTUs)"

energy_filtered <- energy |>
  filter(measurement == energy_type)

usage_cols <- energy_filtered |>
  select(
    -building_address,
    -borough,
    -measurement,
    -postcode,
    -latitude,
    -longitude,
    -community_board,
    -council_district,
    -census_tract,
    -bin,
    -bbl,
    -nta
  ) |>
  colnames()

energy_numeric <- energy_filtered |>
  mutate(across(all_of(usage_cols), ~ suppressWarnings(as.numeric(.)))) |>
  mutate(total_energy = rowSums(across(all_of(usage_cols)), na.rm = TRUE)) |>
  mutate(
    borough_name = factor(
      borough,
      levels = c("1","2","3","4","5"),
      labels = c("Manhattan","Bronx","Brooklyn","Queens","Staten Island")
    )
  )

borough_totals <- energy_numeric |>
  group_by(borough_name) |>
  summarise(
    total_energy_borough = sum(total_energy, na.rm = TRUE)
  )

ggplot(borough_totals) +
  geom_col(aes(
    x = borough_name,
    y = total_energy_borough,
    fill = borough_name
  )) +
  scale_fill_viridis_d(option = "D", name = "Borough Name") + 
  labs(
    title = paste0(energy_type, " by Borough"),
    x = "Borough",
    y = energy_type
  ) +
  theme_minimal() + theme(legend.position = "none")

```

### Plot 3: ___________

```{r}

library(tidyverse)
library(viridis)
library(scales)

wealth_file <- list.files("../data",
                          pattern = "(?i)Neighborhood_Financial_Health.*\\.csv$",
                          full.names = TRUE)
stopifnot(length(wealth_file) >= 1)

wealth_raw <- readr::read_csv(wealth_file[1], show_col_types = FALSE)

wealth <- wealth_raw |>
  rename_with(~ gsub("[^A-Za-z0-9]+", "_", .x)) |>
  rename_with(~ tolower(.x))

if (!"borough" %in% names(wealth)) {
  stop("No 'borough' column in the wealth CSV. Columns are: ",
       paste(names(wealth), collapse = ", "))
}

income_col <- c("median_income", "median_household_income", "income", "medianincome")
have_income <- income_col[income_col %in% names(wealth)]
if (length(have_income) == 0) {
  stop("No income column found. Looked for: ",
       paste(income_col, collapse = ", "), 
       ". Columns are: ", paste(names(wealth), collapse = ", "))
}

wealth_boro <- wealth |>
  mutate(
    borough = stringr::str_trim(borough),
    income_val = suppressWarnings(as.numeric(.data[[have_income[1]]]))
  ) |>
  filter(!is.na(borough), !is.na(income_val)) |>
  group_by(borough) |>
  summarise(median_income = median(income_val, na.rm = TRUE), .groups = "drop")

compare_big <- per_capita |>
  select(borough, total_energy, total_students, energy_per_student) |>
  inner_join(wealth_boro, by = "borough") |>
  drop_na(total_energy, total_students, median_income) |>
  mutate(size_pts = scales::rescale(total_students, to = c(4, 12)))

if (!exists("pal")) {
  pal <- setNames(
    viridis(5, option = "D"),
    c("Manhattan","Bronx","Brooklyn","Queens","Staten Island")
  )
}

ggplot(compare_big, aes(median_income, total_energy, color = borough)) +
  geom_point(aes(size = size_pts)) +
  geom_text(aes(label = borough), vjust = -1, size = 4, show.legend = FALSE) +
  scale_size_identity() +
  scale_color_manual(values = pal) +
  scale_y_continuous(labels = label_comma(), expand = expansion(mult = c(0.1, 0.2))) +
  scale_x_continuous(labels = label_comma(), expand = expansion(mult = c(0.1, 0.2))) +
  labs(
    title = "How Borough Wealth Relates to Total School Energy Use",
    x = "Median household income (USD)",
    y = "Total school energy usage (mmBTUs)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```


### Plot 4: ___________ 

